Modular Model-Checking for Fault-Tolerant Distributed System Implementations

### 2016-12-02 doc/om-paper/paper.tex
Updates to section 4, including new example for ASM subsection

\title{Modular Model-Checking for Fault-Tolerant Distributed System Implementations}
\author{Benjamin F Jones and Lee Pike}
\institute{Galois, Inc.}

@@

%% Two examples of model-specific abstractions include the following: (1) Suppose we model a system in which a process that broadcasts the values $\{0, 1\}$ in the nominal case, but when the process suffers a fault, it broadcasts no value. To model the system, it is typical to define an enumerated type of broadcast values as $$\{0, 1, NONE\}$$ However, the enumeration conflates the environment (i.e., faults) and system (i.e., the non-faulty broadcast values). (2) Another example is to model message passing as shared state between processes to reduce the number of state variables in the model, since the message passing mechanism is elided. While these optimizations might be safe under particular assumptions, they require careful meta-level reasoning and justification and divorce the model from any realization of it.

@@

\subsubsection{The Synchronous Kibitzer}\label{sec:kibitzer}
Synchronous observers can be used to model the environment, and in particular, to compositionally inject faults. The typical approach to injecting faults into a formal model of a distributed system is to add new state variables to each process representing its fault state. Then additional state variables are added to represent the nondeterminism introduced by faults into the state or messages by a process.

@@

Not only is modeling faults expensive, but it is generally non-compositional insofar as the fault model pervades the entire model.

Another difficulty with adding faults to a model is that the number of states that must be introduced may be dependent non-obviously on other aspects of the fault model, specific protocol, and system size. Furthermore, since SAT-based model-checking is NP-complete, it is important not to introduce any state beyond the minimal required. Such constraints lead to ``meta-model'' reasoning, such as the following, as described by Rushby:

\begin{quote}
To achieve the full range of faulty behaviors, it seems that a faulty source should be able to send a \emph{different} incorrect value to each relay, and this requires $n$ different values. It might seem that we need some additional incorrect values so that faulty relays can exhibit their full range of behaviors. It would certainly be safe to introduce additional values for this purpose, but the performance of model checking is very sensitive to the size of the state space, so there is a countervailing argument against introducing additional values. A little thought will show that the way faulty relays have their most significant impact is by tipping the majority vote in a receiver one way or the other, and this can only be achieved if they use the same values as nonfaulty relays. Hence, we decide against further extension to the range of values~\cite{}.
\end{quote}

To address these issues, we introduce a synchronous observer that injects faults that we call a \emph{synchronous kibitzer}. The kibitzer decomposes the state and transitions associated with the fault model from the system. For the sake of concreteness, we introduce a particular fault model, the hybrid fault model of Thambidurai and Park~\cite{}. The fault model distinguishes Byzantine (or arbitrary), symmetric, and manifest faults. This fault model applies to broadcast systems, in which a process is expected to broad the same value to multiple receivers. A \emph{Byzantine fault} is one in which a process that is intended to broadcast the same value to other processes may instead broadcast arbitrary values to different receivers (including no value or the correct value). A \emph{symmetric fault} is one in which a process may broadcast the same, but incorrect, value to other processes, and a \emph{manifest fault} is one in which a process's broadcast fault is detectable by the receivers; e.g., by performing a cyclic redundancy check (CRC) or because the value arrives outside of a predetermined window.

%% In specifying a fault-tolerant system, we must also specify a \emph{maximum fault assumption} (MFA), which specifies the maximum number of of each kind of fault that may be present in the system for some system property to hold.

Define an enumeration of fault types $$\textnormal{FAULT\_TYPE} = \{none, byz, sym, man\}$$ As in the previous section, let $\Id \subset \mathbb{N}$ be a finite set of indices, and let the variable $$faults: \Id \rightarrow \textnormal{FAULT\_TYPE}$$ range over possible mappings from process indices to faults.

@@

With this machinery, we can define the semantics of faults by constraining the relationship between a message broadcast and the values received by the recipients. For a nonfaulty process that broadcasts, every recipient receives the sent message.
\begin{align*}
&nonfaulty\_constraint =\\
  &\quad \forall i, j \in \Id, t \in \mathbb{R}.\\
  &\quad\quad faults(i) = none\\
  &\quad\quad\quad \textnormal{ implies } recv(i, j, t) = send(i, j, t)
\end{align*}
\noindent
For symmetric faults, we there is no requirement that the messages sent are the ones received, only that every recipient receives the same value.
\begin{align*}
&sym\_constraint =\\
  &\quad \forall i, j, k \in \Id, t \in \mathbb{R}.\\
  &\quad\quad (faults(i) = sym \textnormal{ and } broadcast(i, \{j, k\}, t, m))\\
  &\quad\quad\quad \textnormal{ implies } recv(i, j, t) = recv(i, k, t)
\end{align*}
\noindent
Byzantine faults are left completely unconstrained.

@@

% ------------------------------------------------------------
\subsection{Abstract State Machines}\label{sec:abstract}\label{sec:asms}
In model-checking, counterexamples to properties are traces of states. A state is a collection of variable values, and in large models, there are a lot. A single state includes assignments to each of the variables, and the counterexample traces we dealt with (see Section~\ref{sec:experimental}) can contain hundreds of state variables. In practice, counterexamples depend on a small subset of these. The user has to manually separate the state variables that matter from those that do not for a specific property. It is not enough prevent variables that stutter from displaying since some state variables might be updated without affecting the property. The problem is particularly pertinent when model-checking models that have sufficient detail for implementation synthesis, as is the goal in our work.

@@

But the relationship between many of the state variables have an intuitive structure that can be characterized by a state machine imposed over it. The state machine describes discrete modes or configurations. Each mode captures a relation on a subset of the state variables, as well as the set of possible modes that follow.

The idea is an implementation of abstract state machines (ASMs). ASMs are a generalization of finite state machines originally developed by Gurevich to generalize state machines to operate over arbitrary data structures~\cite{asm}. In our context, we require only a restricted form of ASMs known as a \emph{sequential small-step ASM}:

@@

@@

\lee{probably need an example here.}

@@

OMH is designed to preserve \emph{validity} and \emph{agreement}
properties. Validity states that if the general is non-faulty, then every lieutenant outputs the value sent by the general. More formally, let the general broadcast $v \neq ERR$, and let $l_i$ denote the output for lieutenant $i$:
%
\begin{equation}
  \tag{Validity}
    \forall \,i. \quad l_i = v
\end{equation}
%
Agreement states that each lieutenant outputs the same value. Let $l_i, l_j$ denote the outputs of lieutenants $i, j \in L$, respectively:
%
\begin{equation}
  \tag{Agreement}
    \forall \,i, j. \quad l_i = l_j
\end{equation}
%

>>>>

\title{Modular Model-Checking for Fault-Tolerant Distributed System Implementations}

%\author{Benjamin F Jones and Lee Pike}
\author{Benjamin F Jones\inst{1} \and Lee Pike\inst{2}}

\institute{Galois, Inc., Portland OR 97204,\\
\email{bjones@galois.com}
\and
Galois, Inc., Portland OR 97204,\\
\email{leepike@galois.com}}

@@

%% Two examples of model-specific abstractions include the following: (1) Suppose we model a system in which a process that broadcasts the values $\{0, 1\}$ in the nominal case, but when the process suffers a fault, it broadcasts no value. To model the system, it is typical to define an enumerated type of broadcast values as $$\{0, 1, NONE\}$$ However, the enumeration conflates the environment (i.e., faults) and system (i.e., the nonfaulty broadcast values). (2) Another example is to model message passing as shared state between processes to reduce the number of state variables in the model, since the message passing mechanism is elided. While these optimizations might be safe under particular assumptions, they require careful meta-level reasoning and justification and divorce the model from any realization of it.

@@

\subsubsection{The Synchronous Kibitzer}\label{sec:kibitzer}
Synchronous observers can be used to model the environment, and in particular, inject faults in a compositional way. The typical approach to injecting faults into a formal model of a distributed system is to add new state variables to each process representing its fault state. Then additional state variables are added to represent the non-determinism introduced by faults into the state or messages by a process.

@@

Not only is modeling faults generally expensive, but in typical practice it is generally non-compositional insofar as the fault model pervades the entire model.

Another difficulty with adding faults to a model is that the number of states that must be introduced may depend non-obviously on other aspects of the fault model, specific protocol, and system size. Furthermore, since SAT-based model-checking is NP-complete, it is important not to introduce any state beyond the minimal required. Such constraints lead to ``meta-model'' reasoning, such as the following, as described by Rushby:

\begin{quote}
To achieve the full range of faulty behaviors, it seems that a faulty source should be able to send a \emph{different} incorrect value to each relay, and this requires $n$ different values. It might seem that we need some additional incorrect values so that faulty relays can exhibit their full range of behaviors. It would certainly be safe to introduce additional values for this purpose, but the performance of model checking is very sensitive to the size of the state space, so there is a countervailing argument against introducing additional values. A little thought will show that the way faulty relays have their most significant impact is by tipping the majority vote in a receiver one way or the other, and this can only be achieved if they use the same values as nonfaulty relays. Hence, we decide against further extension to the range of values~\cite{Rushby:OM1}.
\end{quote}

To address these issues, we introduce a synchronous observer that injects faults that we call a \emph{synchronous kibitzer}. The kibitzer decomposes the state and transitions associated with the fault model from the system. For the sake of concreteness, we introduce a particular fault model, the hybrid fault model of Thambidurai and Park~\cite{hybrid}. This fault model distinguishes Byzantine (or arbitrary), symmetric, and manifest faults. It  applies to broadcast systems in which a process is expected to broadcast the same value to multiple receivers. A \emph{Byzantine fault} is one in which a process that is intended to broadcast the same value to other processes may instead broadcast arbitrary values to different receivers (including no value or the correct value). A \emph{symmetric fault} is one in which a process may broadcast the same, but incorrect, value to other processes, and a \emph{manifest fault} is one in which a process's broadcast fault is detectable by the receivers; e.g., by performing a cyclic redundancy check (CRC) or because the value arrives outside of a predetermined window.

%% In specifying a fault-tolerant system, we must also specify a \emph{maximum fault assumption} (MFA), which specifies the maximum number of each kind of fault that may be present in the system for some system property to hold.

Define an enumeration of fault types $$\textnormal{FAULT\_TYPE} = \{none, byz, sym, man\}.$$ As in the previous section, let $\Id \subset \mathbb{N}$ be a finite set of process indices, and let the variable $$faults: \Id \rightarrow \textnormal{FAULT\_TYPE}$$ range over possible mappings from processes to faults.

@@

With this machinery, we can define the semantics of faults by constraining the relationship between a message broadcast and the values received by the recipients. For a nonfaulty process that broadcasts, every recipient receives the sent message.
\begin{align*}
&nonfaulty\_constraint =\\
  &\quad \forall i, j \in \Id, t \in \mathbb{R}.\\
  &\quad\quad faults(i) = none\\
  &\quad\quad\quad \textnormal{ implies } recv(i, j, t) = send(i, j, t)
\end{align*}
\noindent
For symmetric faults, there is no requirement that the messages sent are the ones received, only that every recipient receives the same value.
\begin{align*}
&sym\_constraint =\\
  &\quad \forall i, j, k \in \Id, t \in \mathbb{R}.\\
  &\quad\quad (faults(i) = sym \textnormal{ and } broadcast(i, \{j, k\}, t, m))\\
  &\quad\quad\quad \textnormal{ implies } recv(i, j, t) = recv(i, k, t)
\end{align*}
\noindent
Byzantine faults are left completely unconstrained.

@@

% ------------------------------------------------------------
\subsection{Abstract State Machines}\label{sec:abstract}\label{sec:asms}
In model-checking, counterexamples to properties are traces of states. A single state includes assignments to each of the variables, and the counterexample traces we dealt with (see Section~\ref{sec:experimental}) can contain hundreds of state variables. In practice, the essence of a counterexample depends on a small subset of these. The user has to manually separate the state variables that matter from those that do not for a specific property. Moreover, It is not enough to prevent variables that stutter from displaying (a feature that many off-the-shelf model-checkers provide) since some state variables might be updated without affecting the property. The problem is particularly pertinent when formal models have sufficient detail for implementation synthesis, as is the goal in our work.

@@

But the relationship between many of the state variables often has an intuitive structure, one familiar to the designer, that can be characterized by a state machine imposed over it. The state machine describes discrete modes or configurations. Each mode captures a relation on a subset of the state variables, as well as the set of possible modes that follow.

This idea is an implementation of abstract state machines (ASMs). ASMs are a generalization of finite state machines originally developed by Gurevich to generalize state machines to operate over arbitrary data structures~\cite{asm}. In our context, we require only a restricted form of ASMs known as a \emph{sequential small-step ASM}:

@@

\ben{I'm confused by this definition, but I haven't looked at the reference, so maybe it would be clear if I did. My intuitive notion of ASM involves a set of predicates B which *cover* the reachable states and a transition relation over B that abstracts the one over S.}

@@

\begin{example}
Consider a simple distributed system involving 3 processors: $S$, $L_1$, and $L_2$. The purpose of the system is for $S$ to broadcast a message to $L_1$ and $L_2$ after which they perform a computation on the message and update their local state with the result. One might model this system as a transition system whose state is the product of the local processor states along with the state of the environment and any globally important phenomena such as time.

\begin{figure}
\centering
\begin{tikzpicture}[->,>=stealth',shorten >=1pt,auto,node distance=3cm,
  thick,main node/.style={circle,fill=blue!20,draw,
  font=\sffamily\Large\bfseries,minimum size=8mm}]

  \node[main node] (S) at (0,0) {$S$};
  \node[main node] (L1) at (2,1) {$L_1$};
  \node[main node] (L2) at (2,-1) {$L_2$};

  \path[every node/.style={font=\sffamily\small,
          fill=white,inner sep=1pt}]
    (S)  edge (L1)
         edge (L2);

\end{tikzpicture}
\label{fig:exa-trans}
\caption{Simple message passing system}
\end{figure}

Intuitively, there are subsets of the overall state space which are important to understanding the evolution of such a system. For example, let $A_1$ denote the subset of states in which the processor $S$ has not sent a message, let $A_2$ be the subset in which $S$ has broadcast a message, and let $A_3$ be the subset in which $L_1$ and $L_2$ have each received and processed the message. The union $A_1 \cup A_2 \cup A_3$ covers the set of reachable states of the system and defines an ASM that is useful for reasoning about its dynamics and understanding counter-example traces.
\end{example}

@@

OMH is designed to preserve \emph{validity} and \emph{agreement}
properties. Validity states that if the general is nonfaulty, then every lieutenant outputs the value sent by the general. More formally, let the general broadcast $v \neq ERR$, and let $l_i$ denote the output for lieutenant $i$:
%
\begin{equation}
  \tag{Validity}
    \forall \,i. \quad l_i = v
\end{equation}
%
Agreement states that each lieutenant outputs the same value. Let $l_i, l_j$ denote the outputs of lieutenants $i, j \in L$, respectively:
%
\begin{equation}
  \tag{Agreement}
    \forall \,i, j. \quad l_i = l_j
\end{equation}
%
